---
title: "Estimating Heritability"
author: "Caryn Johansen"
date: "2/27/2017"
output: html_document
---

Purpose: This document is mean to to be the workspace for estimating the heritability of traits collected in the Hirsch et al. paper

Citation:

##Heritability

Phenotypic variance is the sum of the genetic component and the environmental component:

$$V_P = V_G + V_E$$

The genetic component is made up of an additive component and a non-additive (or dominance) component:

$$V_P = V_A + V_D + V_E$$

$V_A$ is the component of the phenotypic variability that is transmitted to progeny. We use _heritability_ to measure how much of
the phenotypic variability of a trait in a given population will be inherited by the progeny of a population. Essentially, it is calculating
what proportion of the phenotypic variation is attributed to additive genetic variation:

**Narrow sense heritability** is how much of the phenotypic variability is explained by the addititve
component of the genetic variability.

$$h^2 = \frac{V_A}{V_P}$$
This value will lie between 0 and 1.

This is defined for a given population at a given time.

**Broad sense heritability** is how much of phenotypic variability is explained by the genetic 
variability

$$H^2 = \frac{V_G}{V_P}$$

The value will lie between 0 and 1. The closer to one, the more of the phenotype variability is
explained by the genetic variability.

## Estimating additive variability

estimatine the variance of the kinship matrix is $V_A$ (citation: my notes from last week)

rrblup

gtca - includes partitioning heritability across markers



## Estimating the phenotypic variability

Load lmer package
```{r}
library(lme4)
library(ggplot2)
library(lsmeans)
```

Load the phenotypic data

```{r, eval=F}
pheno <- read.table("data/processed/WIDIV_2010_paper_phenotypic_data_cleaned.csv", sep=",", header=T, na.strings = ".")

colnames(pheno) <- c("Year", "Rep", "Genotype", colnames(pheno[,4:20]))

str(pheno)
```

Include the population group data

```{r, eval=F}
pop <- read.table("data/processed/Supplemental_Table_1_final_c.csv", sep=",", header=T, na.strings = ".")
pop <- pop[,c(2,5,6)]
pop$Genotype <- as.character(pop$Genotype)
pop$Population.Group <- as.factor(as.integer(pop$Population.Group))

table(pop$Population.Group)
levels(pop$Group)
table(pop$Group, pop$Population.Group)

ggplot(pop, aes(factor(Group))) + geom_bar() # + geom_jitter(aes(Group))
```

Add population information to phenotype data

```{r, eval=F}
pheno <- merge(pheno, pop)

pheno.plot <- pheno[pheno$Year == 2008 & pheno$Rep == 1,]

pheno.plot$Year <- as.factor(as.integer(pheno.plot$Year))
ggplot(pheno.plot, aes(factor(Group))) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Population Group") + ylab("Count") + ggtitle("Phenotype Data Maize Population Groups")
ggsave(filename = "output/phenotype_popStructure_bar.png",width = 12, height = 8)
  
ggplot(pheno.plot, aes(factor(Population.Group))) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


List the phenotypes:

```{r, eval=F}
traits <- colnames(pheno[,c(7:20)])
traits
```

#### Save data space

```{r, eval=F}
save.image("Heritability_Frequency.RData")
```

#### Begin here

```{r}
load("Heritability_Frequency.RData")

table(pop$Population.Group)

table(pop$Population.Group, pop$Group)
```


Wisconsin Diversity Panel Phenotypes

| Trait | Explanation/unit |
| ----- | ---------------- |
| GDD | days to flowering (growing degree days) |
| kernel300 | weight of 300 kernels (g) |
| stalk | stalk diameter (cm) |
|plant_height | cm |
|ear_height | cm |
| leaf_number | number of leaves |
| stover_yield | $t ha^{-1}$|
| juvenile | last leaf with epicuticular wax |
| ear_internode | uppermost internode with developed ear (unit?) |
| internode_length | cm |
|internode\_above_ear | number of internodes above uppermost ear |
| adult | number of leaves with no epicuticular wax |
|percent_juvenile | percent of leave with epicuticular wax |
|percentage_adult | percent of leaves with no epicuticular wax |


### GDD

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=GDD)) +
  geom_boxplot(aes(colour=as.factor(Rep))) +
  geom_jitter(aes(colour=as.factor(Population.Group)))

gdd.lmer <- lmer(GDD ~ (1|Genotype) + Year, data=pheno)
gdd.lm <- lm(GDD ~ Genotype + Year, data=pheno)
gdd.lmer

gdd.re <- ranef(gdd.lmer)

line_means = as.data.frame(summary(lsmeans(gdd.lm,~Genotype)))
plot(gdd.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(gdd.lmer)

vars = as.data.frame(VarCorr(gdd.lmer))
vars
gdd.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
gdd.herit

heritability_traits <- data.frame("trait"="GDD", "H^2"=gdd.herit)

```


### Stalk

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=stalk)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Stalk")
#ggplot(pheno.plot, aes(x=Year, y=leaf_number, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = stalk)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group)))

st.lmer <- lmer(stalk ~ (1|Genotype) + Rep:Year, data=pheno)
st.lm <- lm(stalk ~ Genotype + Rep:Year, data=pheno)
st.lmer

st.re <- ranef(st.lmer)

line_means = as.data.frame(summary(lsmeans(st.lm,~Genotype)))
plot(st.re$Genotype[,1],line_means$lsmean);abline(0,1)

par(mfrow=c(3,1))
hist(st.re$Genotype[,1])
hist(line_means$lsmean)
hist(pheno$stalk)

summary(st.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(st.lmer))
vars

st.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
st.herit

row <- data.frame("trait" ="stalk", "H^2"= st.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Plant Height

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=plant_height)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Plant Height")
#ggplot(pheno.plot, aes(x=Year, y=plant_height, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = plant_height)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group)))

ph.lmer <- lmer(plant_height ~ (1|Genotype) + Rep:Year, data=pheno)
ph.lmer
summary(ph.lmer)

ph.lm <- lm(plant_height ~ Genotype + Rep:Year, data=pheno)
ph.lmer

ph.re <- ranef(ph.lmer)


line_means = as.data.frame(summary(lsmeans(ph.lm,~Genotype)))
plot(ph.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(ph.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(ph.lmer))
vars

ph.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
ph.herit

row <- data.frame("trait" ="plant_height", "H^2"= ph.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Ear Height

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=ear_height)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Ear Height")
#ggplot(pheno.plot, aes(x=Year, y=plant_height, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = ear_height)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Ear Height")

eh.lmer <- lmer(ear_height ~ (1|Genotype) + Rep:Year, data=pheno)
eh.lmer
summary(eh.lmer)

eh.lm <- lm(ear_height ~ Genotype + Rep:Year, data=pheno)
eh.lmer

eh.re <- ranef(eh.lmer)


line_means = as.data.frame(summary(lsmeans(eh.lm,~Genotype)))
plot(eh.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(eh.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(eh.lmer))
vars

eh.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
eh.herit

row <- data.frame("trait" ="ear_height", "H^2"= eh.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Leaf Number

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=leaf_number)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Leaf Number")
#ggplot(pheno.plot, aes(x=Year, y=leaf_number, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = leaf_number)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Leaf Number")

ln.lmer <- lmer(leaf_number ~ (1|Genotype) + Rep:Year, data=pheno)
ln.lmer
summary(ln.lmer)

ln.lm <- lm(leaf_number ~ Genotype + Rep:Year, data=pheno)
ln.lmer

ln.re <- ranef(ln.lmer)


line_means = as.data.frame(summary(lsmeans(ln.lm,~Genotype)))
plot(ln.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(ln.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(ln.lmer))
vars

ln.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
ln.herit

row <- data.frame("trait" ="leaf_number", "H^2"= ln.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Stover Yield

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=stover_yield)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Stover Yield")
#ggplot(pheno.plot, aes(x=Year, y=stover_yield, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = stover_yield)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Stover Yield")

sy.lmer <- lmer(stover_yield ~ (1|Genotype) + Rep:Year, data=pheno)
sy.lmer
summary(sy.lmer)

sy.lm <- lm(stover_yield ~ Genotype + Rep:Year, data=pheno)
sy.lmer

sy.re <- ranef(sy.lmer)


line_means = as.data.frame(summary(lsmeans(sy.lm,~Genotype)))
plot(sy.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(sy.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(sy.lmer))
vars

sy.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
sy.herit

row <- data.frame("trait" ="stover_yield", "H^2"= sy.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Juvenile

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=juvenile)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Juvenile")
#ggplot(pheno.plot, aes(x=Year, y=juvenile, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = juvenile)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Juvenile")

jvlmer <- lmer(juvenile ~ (1|Genotype) + Rep:Year, data=pheno)
jvlmer
summary(jvlmer)

jvlm <- lm(juvenile ~ Genotype + Rep:Year, data=pheno)
jvlmer

jvre <- ranef(jvlmer)


line_means = as.data.frame(summary(lsmeans(jvlm,~Genotype)))
plot(jvre$Genotype[,1],line_means$lsmean);abline(0,1)
summary(jvlmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(jvlmer))
vars

jvherit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
jvherit

row <- data.frame("trait" ="juvenile", "H^2"= jvherit)
heritability_traits <- rbind(heritability_traits, row)
```

### Percent Juvenile

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=percent_juvenile)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Percent Juvenile")
#ggplot(pheno.plot, aes(x=Year, y=percent_juvenile, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = percent_juvenile)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Percent Juvenile")

pj.lmer <- lmer(percent_juvenile ~ (1|Genotype) + Rep:Year, data=pheno)
pj.lmer
summary(pj.lmer)

pj.lm <- lm(percent_juvenile ~ Genotype + Rep:Year, data=pheno)
pj.lmer

pj.re <- ranef(pj.lmer)

line_means = as.data.frame(summary(lsmeans(pj.lm,~Genotype)))
plot(pj.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(pj.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(pj.lmer))
vars

pj.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
pj.herit

row <- data.frame("trait" ="percent_juvenile", "H^2"= pj.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Ear Internode

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=ear_internode)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Ear Internode")
#ggplot(pheno.plot, aes(x=Year, y=ear_internode, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = ear_internode)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Ear Internode")

ei.lmer <- lmer(ear_internode ~ (1|Genotype) + Rep:Year, data=pheno)
ei.lmer
summary(ei.lmer)

ei.lm <- lm(ear_internode ~ Genotype + Rep:Year, data=pheno)
ei.lmer

ei.re <- ranef(ei.lmer)


line_means = as.data.frame(summary(lsmeans(ei.lm,~Genotype)))
plot(ei.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(ei.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(ei.lmer))
vars

ei.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
ei.herit

row <- data.frame("trait" ="ear_internode", "H^2"= ei.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Internode Length

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=internode_length)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Internode Length")
#ggplot(pheno.plot, aes(x=Year, y=internode_length, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = internode_length)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Internode Length")

il.lmer <- lmer(internode_length ~ (1|Genotype) + Rep:Year, data=pheno)
il.lmer
summary(il.lmer)

il.lm <- lm(internode_length ~ Genotype + Rep:Year, data=pheno)
il.lmer

il.re <- ranef(il.lmer)


line_means = as.data.frame(summary(lsmeans(il.lm,~Genotype)))
plot(il.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(il.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(il.lmer))
vars

il.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
il.herit

row <- data.frame("trait" ="internode_length", "H^2"= il.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Internode above Ear

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=internode_above_ear)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Internode Above Ear")
#ggplot(pheno.plot, aes(x=Year, y=internode_above_ear, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = internode_above_ear)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Internode Above Ear")

iae.lmer <- lmer(internode_above_ear ~ (1|Genotype) + Rep:Year, data=pheno)
iae.lmer
summary(iae.lmer)

iae.lm <- lm(internode_above_ear ~ Genotype + Rep:Year, data=pheno)
iae.lmer

iae.re <- ranef(iae.lmer)


line_means = as.data.frame(summary(lsmeans(iae.lm,~Genotype)))
plot(iae.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(iae.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(iae.lmer))
vars

iae.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
iae.herit

row <- data.frame("trait" ="internode_above_ear", "H^2"= iae.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Adult

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=adult)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Adult")
#ggplot(pheno.plot, aes(x=Year, y=adult, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = adult)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Adult")

ad.lmer <- lmer(adult ~ (1|Genotype) + Rep:Year, data=pheno)
ad.lmer
summary(ad.lmer)

ad.lm <- lm(adult ~ Genotype + Rep:Year, data=pheno)
ad.lmer

ad.re <- ranef(ad.lmer)


line_means = as.data.frame(summary(lsmeans(ad.lm,~Genotype)))
plot(ad.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(ad.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(ad.lmer))
vars

ad.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
ad.herit

row <- data.frame("trait" ="adult", "H^2"= ad.herit)
heritability_traits <- rbind(heritability_traits, row)
```

### Percent Adult

```{r, fig.show='hold', fig.width=12, fig.height=8}
ggplot(pheno.plot, aes(x=Year, y=percentage_adult)) +
  geom_boxplot(aes(color=as.factor(Rep))) +
  labs(title = "Percent Adult")
#ggplot(pheno.plot, aes(x=Year, y=percentage_adult, colour=as.factor(Entry))) +
#  geom_boxplot() +
#  theme(legend.position="none")

ggplot(pheno.plot, aes(x=Population.Group, y = percentage_adult)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(Population.Group))) +
  labs(title = "Percent Adult")

pa.lmer <- lmer(percentage_adult ~ (1|Genotype) + Rep:Year, data=pheno)
pa.lmer
summary(pa.lmer)

pa.lm <- lm(percentage_adult ~ Genotype + Rep:Year, data=pheno)
pa.lmer

pa.re <- ranef(pa.lmer)


line_means = as.data.frame(summary(lsmeans(pa.lm,~Genotype)))
plot(pa.re$Genotype[,1],line_means$lsmean);abline(0,1)
summary(pa.lmer)
#calculate the estimated variances, std dev, correlations between random-effects terms in the 
# mixed effect model
vars = as.data.frame(VarCorr(pa.lmer))
vars

pa.herit <- vars[1,]$vcov/(vars[1,]$vcov + vars[2,]$vcov)
pa.herit

row <- data.frame("trait" ="percentage_adult", "H^2"= pa.herit)
heritability_traits <- rbind(heritability_traits, row)
```

Heritability table:

```{r}
heritability_traits
```

