---
title: "Imputing Phenotype Data"
author: "Caryn Johansen"
date: "4/18/2017"
output: html_document
---

modified script from Rongkui

```{r}
#setwd("~/Dropbox/UCD/rotations/RRI/maize_genomics_project/")
library(reshape2)
library(ggplot2)
## Loading phenotype data
td <- read.table("data/processed/WIDIV_2010_paper_phenotypic_data_cleaned.csv",sep=",", header=T, na.strings = ".")

## model for year effect (line~year*rep(use 1,2 in year1 and 3,4 in year2), with lsmeans function in package lsmeans) instead of removing na's. 
trait2008 <- td[td$Year=="2008",]
trait2009 <- td[td$Year=="2009",]
trait2009$Rep[trait2009$Rep==1] <- 3
trait2009$Rep[trait2009$Rep==2] <- 4
td <- rbind(trait2008, trait2009)

td$Year <- as.factor(as.integer(td$Year))
td$Rep <- as.factor(as.numeric(td$Rep))
td$Entry <- as.factor(as.character(td$Entry))
str(td)
```

```{r}
image(as.matrix(td[,7:20]))
```


```{r}
td2 <- td[,-c(3:6)]

td2_m <- melt(td2, id.vars = c("Year","Rep"))
td2_m$Rep <- as.factor(as.numeric(td2_m$Rep))
td2_m$Year <- as.factor(as.integer(td2_m$Year))
ggplot(td2_m, aes(x=Year, y=value, colour=Rep)) +
  geom_boxplot() +
  facet_wrap(~variable, scale = "free")
```


That there are some lines that always have NA's is a problem, and we have to remove them:

```{r}
to.clean <- data.frame(line=NA, i=NA)

for (i in 1:nrow(td)) {
  # find lines of maize that continuously return only NAs
  if (sum(is.na(td[i,7:20])) == 14) {
    line <- as.character(td[i,3])
    line.idx <- cbind(line, i)
    to.clean <- rbind(to.clean, line.idx)
  }
}

to.clean<- to.clean[-1,]

to.clean$line <- as.factor(as.character(to.clean$line))
to.clean$i <- as.integer(as.character(to.clean$i))

##remove the repeat offender lines
x <- table(to.clean$line)

#only want to remove the repeating lines, so that have na's in all 4 measurment windows (Year by Rep)
to_remove_table <- x[x >= 4]
length(to_remove_table)
to_remove <- to.clean[to.clean$line %in% names(to_remove_table),]

td.clean <- td[-to_remove$i,]
image(as.matrix(td.clean[,7:20]))
```


remove kernal300 as trait

```{r}
td.clean <- td.clean[,-8]
image(as.matrix(td.clean[,7:19]))
```

Calculate least squares means with lsmeans

```{r}
library(lsmeans)

lsmeans.td = as.data.frame(td.clean[unique(td.clean$Entry),]$Entry)
#there are 631 maize lines now
td.clean.lines <- as.character(unique(td.clean$Entry))

colnames(lsmeans.td) <- "Entry"
pred.td <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])

imputed_pheno_list <- list()

```

For each trait, fit a linear model, then use lsmeans

```{r}
traits <- colnames(td.clean)[7:19]
traits
```

It won't work as a loop, because not each trait has more than one rep and year (some traits have only one rep, or only one year)

GDD
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
t <- colnames(td.clean)[7]
t

#raw trait values
y <- td.clean[,7]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year, data = td.clean, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 40 lines that had na's and were omitted in the lm stage
missing_lines <- td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))]
```

Stalk
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
t <- colnames(td.clean)[8]
t

#raw trait values
y <- td.clean[,8]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Rep, data = td.clean, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")
pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Plant Height
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#In order to use the interaction between Rep and Year, the reps here  need to be 1 and 2, across both years (not 1,2,3,4)
td.year2008 <- td.clean[td.clean$Year == "2008",]
td.year2009 <- td.clean[td.clean$Year == "2009",]
td.year2009$Rep[td.year2009$Rep == 3] <- 1
td.year2009$Rep[td.year2009$Rep == 4] <- 2

td.clean.ph <- rbind(td.year2008, td.year2009)

t <- colnames(td.clean)[9]
t

#raw trait values
y <- td.clean[,9]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Ear height
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#same here
t <- colnames(td.clean)[10]
t

#raw trait values
y <- td.clean[,10]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Leaf Number
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#same here
t <- colnames(td.clean)[11]
t

#raw trait values
y <- td.clean[,11]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Stover Yield
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean here - only collected in year 2008, so use Rep as the fixed effect
t <- colnames(td.clean)[12]
t

#raw trait values
y <- td.clean[,12]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Rep, data = td.clean, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait
#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```


Precent Juvenile
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - four reps across two years
t <- colnames(td.clean)[13]
t

#raw trait values
y <- td.clean[,13]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 97 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Ear Internode -SKIP
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - two reps in one year
t <- colnames(td.clean)[14]
t

#raw trait values
y <- td.clean[,14]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Rep, data = td.clean, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#DO I care about ear internode? This lost me 50 maize lines...
#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Internode Length
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - four reps across two years
t <- colnames(td.clean)[15]
t

#raw trait values
y <- td.clean[,15]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Internode above ear length - SKIP
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean here - two reps in one year (2009)
t <- colnames(td.clean)[16]
t

#raw trait values
y <- td.clean[,16]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Rep, data = td.clean, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Adult
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - four reps, two years
t <- colnames(td.clean)[17]
t

#raw trait values
y <- td.clean[,17]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Percent Juvenile
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - four reps, two years
t <- colnames(td.clean)[18]
t

#raw trait values
y <- td.clean[,18]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Percentage Adult
```{r}
pred.trait <- as.data.frame(td.clean[, c("Year", "Rep", "Entry")])
#use td.clean.ph here - four reps, two years
t <- colnames(td.clean)[19]
t

#raw trait values
y <- td.clean[,19]

#trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean, na.action = na.omit)
trait.lm <- lm(y ~ 0 + Entry + Year*Rep, data = td.clean.ph, na.action = na.omit)

pred.test <- lsmeans(trait.lm, "Entry")

pred.test.t = summary(pred.test)[,c(1:2)]
colnames(pred.test.t)[2] <- t

pred.trait = merge(pred.trait,pred.test.t,by = "Entry")
pred.trait <- pred.trait[,c(1,4)]
pred.trait <- pred.trait[!duplicated(pred.trait),]

imputed_pheno_list[[t]] <- pred.trait

#there are 141 lines that had na's and were omitted in the lm stage
missing_lines <- c(missing_lines, td.clean.lines[!(td.clean.lines %in% as.character(pred.test.t$Entry))])
missing_lines <- unique(missing_lines)
```

Remove duplicated lines

```{r}
#just entry and imputed traits
#td.pred <- pred.td[,c(1,4:14)]
#td.predicted <- td.pred[!duplicated(td.pred),]

#write.table(td.predicted, "~/Dropbox/UCD/rotations/RRI/maize_genomics_project/data/processed/traits_imputed_04182017.txt", sep=",", quote = F, row.names = F)
```

Save list

```{r}
#save(imputed_pheno_list, file = "imputed_phenotypes_list")
```

