---
title: "WGCNA output"
author: "Caryn Johansen"
date: "2/16/2017"
output: html_document
---


Purpose: to become familiar with Rongkui's WGCNA output data

```{r}
setwd("~/Dropbox/UCD/rotations/RRI/maize_genomics_project/")

library(gplots)
library(ggplot2)
library(qqman)
```

Load wgcna Rdata file

load expression data

```{r}
#load("data/wgcna_gwas_output/maize490-networkConstruction-stepByStep.RData")
#load("data/wgcna_gwas_output/503line_names.RData")
load("data/wgcna_gwas_output/line_names_361.RData")
load("data/wgcna_gwas_output/maize361-networkConstruction-stepByStep.RData")

#exp <- read.table("../Hirsch_correlations/data/processed/maize_503genotypes_FPKM_expressed.txt", header=T)
#write.table(expdata503_names, file="maize_503Exp_originalNames.txt", sep="\n", quote=F, row.names = F)

genotype <- read.table("data/ZeaGBSv27_s_vF.bimbam", sep=",")
```

#Data processing

Maize Genotypes were downloaded from panzea genotype data. I used the Maize GBS genotypes version 2.7 (http://cbsusrv04.tc.cornell.edu/users/panzea/download.aspx?filegroupid=4)

Specifically, the file ZeaGBSv27_publicSamples_imputedV5_AGPv3_20170206.h5, imputed genotypes with ACPv3 coordinates.

To process this file, this was first converted into hapmap format, then converted into VCF [make one step?]. In addition, the panzea
data has genotype data for 17,280 public maize lines. The genotype data was subsetted with a list of maize lines used for the 
Wisconsin diversity panel. In addition, only maize lines with complete expression were used, and, for the final
GWAS, lines with high frequency of missing genotypes were filtered out. This resulted in 304 maize lines included
in the analysis.

# Filtering

Used compared vcftools to tassel minor allele frequency filtering:

Missing data by locus:

```{r, eval=F}
vcf_lmiss <- read.table("data/wgcna_gwas_output/ZeaGBSv27_sF_vcfInfo.lmiss", header = T)
tassel_lmiss <- read.table("data/wgcna_gwas_output/ZeaGBSv27_stF_vcfInfo.lmiss", header=T)

dim(vcf_lmiss)
dim(tassel_lmiss)

par(mfrow=c(1,2))
hist(vcf_lmiss$F_MISS, main="vcftool filtered")
hist(tassel_lmiss$F_MISS, main="tassel filtered")
```

By locus, vcftools filtered more than tassel did.

Missing data by locus:
```{r, eval=F}
vcf_imiss <- read.table("data/wgcna_gwas_output/ZeaGBSv27_sF_vcfInfo.imiss", header = T)
tassel_imiss <- read.table("data/wgcna_gwas_output/ZeaGBSv27_stF_vcfInfo.imiss", header=T)

par(mfrow=c(1,2))
hist(vcf_imiss$F_MISS, breaks = 20)
hist(tassel_imiss$F_MISS, breaks = 20)

bad_lines <- vcf_imiss[vcf_imiss$F_MISS >= 0.4,]
bad_lines <- as.character(bad_lines$INDV)
#load in the lines used for the analysis, remove poor-quality lines, and re-run GWAS
maize304 <- read.table("data/maize_304_GBS.txt")
taxa <- as.character(maize304$V1)

taxa <- taxa[!(taxa %in% bad_lines)]
#write.table(taxa,"data/maize_304_GBS_c.txt", quote = F, row.names = F, col.names = F)
```

Run GWAS without the maize lines with a high frequency of missing data. Remember to do this first next time.


##Site Frequency Spectrum

```{r, fig.show='hold', fig.width=12, fig.height=8, eval=F}
library(reshape2)
library(zoo)
a_freq <- read.table("ZEAGBSv27_s_F_freq.csv", sep=",", header=T)
a_freq <- a_freq[,-c(9:16)]

af <- a_freq[,c(1,2,6,8)]

afl <- reshape(af, varying=c("A1_FREQ","A2_FREQ"), 
               idvar=c("CHROM", "POS"), v.names="FREQ", times=c("A1_FREQ", "A2_FREQ"), direction = "long")
afl$time <- as.factor(afl$time)

afl_chr1_A1 <- afl[afl$time == "A1_FREQ" & afl$CHROM == 1,]
afl_chr1_A2 <- afl[afl$time == "A2_FREQ" & afl$CHROM == 1,]

afl_chr1_A1_roll <- rollmean(afl_chr1_A1$FREQ,k=1000)
afl_chr1_A2_roll <- rollmean(afl_chr1_A2$FREQ, k=1000)

afl_chr1 <- data.frame(POS=seq(1:length(afl_chr1_A1_roll)), ALLELE=rep("A1", length(afl_chr1_A1_roll)), ROLL_FREQ=afl_chr1_A1_roll)

a2_chr <- data.frame(POS=seq(1:length(afl_chr1_A2_roll)),ALLELE=rep("A2", length(afl_chr1_A2_roll)), ROLL_FREQ=afl_chr1_A2_roll)

afl_chr1 <- rbind(afl_chr1, a2_chr)

#afl_roll <- cbind(afl_A1_roll, afl_A2_roll)

ggplot(afl_chr1, aes(x=POS, y=ROLL_FREQ, colour=as.factor(ALLELE)) ) +
  geom_line()

ggplot(afl, aes(x=POS, y=FREQ, colour=as.factor(time))) + geom_line()

```


```{r,, eval=F}
hist(a_freq$A1_FREQ, main = "Major Allele Frequency")
min(a_freq$A1_FREQ)
a_freq[a_freq$A1_FREQ == min(a_freq$A1_FREQ),]
```

Minor Allele Frequency

```{r, eval=F}
hist(a_freq$A2_FREQ, main="Minor Allele Frequency", freq=T)
ggplot(a_freq, aes(A2_FREQ)) + geom_density(aes(y=..scaled..)) + 
  labs(title = "Minor Allele Site Frequency Spectrum", x= "Minor Allele Frequency", y = "Density of AF")
```


######
##WGCNA Module GWAS

GWA done for each module PC1, using genotype data for included lines

$$y = x\beta + u + \epsilon$$

Where:
$y$ is the n-vector WGCNA module PC1  
$x$ is n-vector of marker genotypes  
$\beta$ is the effect size of the marker  
$u$ is the vector of random effects $u \sim MVN_n(0, \lambda\tau^{-1} K)$  
$\epsilon$ is the error vector $\epsilon \sim MVN_n(0, \tau^{-1} I_n)$

```{r}
mods <- colnames(MEs_361)
mods
```

```{r, fig.show='hold', fig.width=12, fig.height=8}

for (i in 1:length(mods)) {
  mod_path <- paste(paste("data/wgcna_gwas_output/", mods[i], sep=""), "/output/", sep="")
  assoc.file <- paste(paste("ZeaGBSv27_", mods[i], sep=""), ".assoc.txt", sep="")
  rel.file <- "ZeaGSv27_sF.sXX.txt"
  ME.assoc <- read.table(paste(mod_path, assoc.file, sep=""), header=T)

  ME.assoc$chr <- as.numeric(as.integer(ME.assoc$chr))
  
  #par(mfrow=c(1,2), las=1)
  manhattan(ME.assoc, chr="chr", snp="rs", bp="ps", p="p_score", main=mods[i])
  qq(ME.assoc$p_score, main=paste(mods[i], "Q-Q plot of GWAS p-values", sep="\n"))
}
```

Grey60 Module

```{r, fig.show='hold', fig.width=12, fig.height=8}
greyM <- mods[1]
greyM

grey.assoc <- read.table("data/wgcna_gwas_output/MEgrey60/output/ZeaGBSv27_MEgrey60.assoc.txt", header=T)

grey.assoc$chr <- as.numeric(as.integer(grey.assoc$chr))

grey_high <- grey.assoc[grey.assoc$p_score < 0.0001,]
soi <- grey_high$rs

manhattan(grey.assoc, chr="chr", snp="rs", bp="ps", p="p_score", main=greyM)#, highlight = soi)
manhattan(subset(grey.assoc, chr == 10), chr="chr", snp="rs", bp="ps", p="p_score", main=greyM)
qq(grey.assoc$p_score, main=paste(greyM, "Q-Q plot of GWAS p-values", sep="\n"))
```

Blue Module

```{r, fig.show='hold', fig.width=12, fig.height=8}
blueM <- mods[2]

blue.assoc <- read.table("data/wgcna_gwas_output/MEblue/output/ZeaGBSv27_MEblue.assoc.txt", header=T)

blue.assoc$chr <- as.numeric(as.integer(blue.assoc$chr))

blue_high <- blue.assoc[blue.assoc$p_score < 0.0001,]
soi <- blue_high$rs

manhattan(blue.assoc, chr="chr", snp="rs", bp="ps", p="p_score", main=blueM, highlight = soi)
manhattan(subset(blue.assoc, chr == 10), chr="chr", snp="rs", bp="ps", p="p_score", main=blueM)
qq(blue.assoc$p_score, main=paste(blueM, "Q-Q plot of GWAS p-values", sep="\n"))
```

###Midnight Blue Module

```{r, fig.show='hold', fig.width=12, fig.height=8}
MMM <- mods[4]
MMM

MMM.assoc <- read.table("data/wgcna_gwas_output/MEmidnightblue/output/ZeaGBSv27_MEmidnightblue.assoc.txt", header=T)

MMM.assoc$chr <- as.numeric(as.integer(MMM.assoc$chr))

MMM_high <- MMM.assoc[MMM.assoc$p_score < 0.0001,]
soi <- MMM_high$rs
MMM7 <- MMM.assoc[MMM.assoc$chr == 7,]
grmzm2g001114 <- MMM7[MMM7$ps >= 82620000,]

manhattan(MMM.assoc, chr="chr", snp="rs", bp="ps", p="p_score", main=MMM, highlight = soi)
manhattan(subset(MMM.assoc, chr == 7), chr="chr", snp="rs", bp="ps", p="p_score", main=MMM, highlight = soi)
qq(MMM.assoc$p_score, main=paste(MMM, "Q-Q plot of GWAS p-values", sep="\n"))
```

The significant peak on Chromosome 7 is GRMZM2G462325 (19083049 - 19101243). It is involved in a protein processing pathway in the ER (pathway zma04141). It is dolichyl-diphosphooligosaccharide--protein glycosyltransferase

```{r, fig.show='hold', fig.width=12, fig.height=8}
GRMZM2G462325 <- MMM7[MMM7$ps > 19083049 & MMM7$ps < 19101243, ]
gene <- as.character(GRMZM2G462325$rs)

manhattan(subset(MMM.assoc, chr == 7), chr="chr", snp="rs", bp="ps", p="p_score", main=MMM, highlight = gene)
```

alleles at peak:

```{r}
qtl_af <- genotype[genotype$V1 == "S7_19083746",]
```


###Turquoise

```{r, fig.show='hold', fig.width=12, fig.height=8}
turq <- mods[8]
turq

turq.assoc <- read.table("data/wgcna_gwas_output/MEturquoise/output/ZeaGBSv27_MEturquoise.assoc.txt", header=T)

turq.assoc$chr <- as.numeric(as.integer(turq.assoc$chr))

turq_high <- turq.assoc[turq.assoc$p_score < 0.0001,]
soi <- turq_high$rs

manhattan(turq.assoc, chr="chr", snp="rs", bp="ps", p="p_score", main=turq, highlight = soi)
#manhattan(subset(turq.assoc, chr == 7), chr="chr", snp="rs", bp="ps", p="p_score", main=turq, highlight = soi)
qq(turq.assoc$p_score, main=paste(turq, "Q-Q plot of GWAS p-values", sep="\n"))
```

Peaks in Chr 3: 
GRMZM2G409514 (222004484 - 222004684)
AC210422.3_FG011 (222004584 - )
GRMZM2G011608 (start - end: 230740989 - 230747356)

```{r}
turq3 <- turq.assoc[turq.assoc$chr == 3,]
GRMZM2G011608 <- turq3[turq3$ps >= 230740989 & turq3$ps <= 230747356, ]

GRMZM2G409514 <- turq3[turq3$ps >= 222004484 & turq3$ps <= 222004684,]

soi <- c(as.character(GRMZM2G011608$rs), as.character(GRMZM2G409514$rs))
soi

manhattan(turq3, chr="chr", snp="rs", bp="ps", p="p_score", main=turq, highlight = soi)
```



# Modules and their Genes

